---
title: "extra"
author: "technicalelvis.com"
date: "7/21/2019"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(ggplot2)
library(knitr)
library(ggplot2)
library(gridExtra)
library(mvtnorm)
```

## Hypergeometric

Consider a playing cards example with _a_ aces and _b_ non-aces (ie. the 48 other cards that are not aces). We draw _n_ cards out of the deck at random without replacement, such that 
${a + b \choose n}$ cards are equally likely. Let _X_ be the number of aces in the in your hand after drawing _n_ cards. _X_ is said to have the
_Hypergeometric distribution_ with parameters a, b and n; we denote this by _X ~ HGeom(a,b,n)_.

### Hypergeometric Density

The plot shows the probability of having *exactly* 1, 2, 3, or 4 aces in a 5 card hand from a deck with 52 cards.

```{r dhyper, echo=F, fig.width=6, fig.height=4, fig.align = "center"}

chap3.3 <- function() {
  # Prob x=1:4 where X ~ HyperGeom(4, 48, 5)
  # from a deck with 
  # - 1:4 aces
  # - 48 non aces
  # - choose 5 cards per hand
  x <- 1:4
  y <- dhyper(x, 4, 48, 5)
  df <- data.frame(x=x, y=y)
  g <- ggplot(df, aes(x=x, y=y)) +
    geom_point() +
    geom_linerange(aes(x=x, ymax=y, ymin=0)) +
    scale_x_discrete(name="x", limits=x) +
    ylab("y") +
    theme(plot.title = element_text(hjust = 0.5)) +
    labs(title="Hypergeometric Density")
  g
}

chap3.3()
```

### Hypergeometric Cumulative Distribution

The plot shows the probability of having more than _x_ aces in a 5 card hand from a deck containing 52 cards.

```{r phyper, echo=F, fig.width=6, fig.height=4, fig.align = "center"}

chap3.4 <- function() {
  # Prob x=1:4 where X ~ HyperGeom(4, 48, 5)
  # from a deck with 
  # - 1:4 aces
  # - 48 non aces
  # - choose 5 cards per hand
  x <- 1:4
  y <- phyper(x, 5, 48, 5, lower.tail = F)
  df <- data.frame(x=x, y=y)
  g <- ggplot(df, aes(x=x, y=y)) +
    geom_line() +
    scale_x_discrete(name="x", limits=x) +
    ylab("y") +
    theme(plot.title = element_text(hjust = 0.5)) +
    labs(title="Hypergeometric Cumulative Distribution")
  g
}

chap3.4()
```


## Uniform

A uniform r.v. on the interval (a,b) is a completely random number between _a_ and _b_. The PDF is _constant_ over the interval. A common example of a uniform distribution is a [dice roll](http://rolladie.net/).

### Uniform Density

Suppose that a die is rolled and $X$ is the number face up, what the probability that the value is 1, 2, 3, 4, 5, or 6.

```{r dunif, echo=F, fig.width=6, fig.height=4, fig.align="center"}
do_dunif <- function() {
  # PDF
  # x <- seq(-0.5, 1.5, 1/10000)
  x <- seq(1, 6, 1)
  y <- dunif(x, 1, 6)
  datums <- data.frame(x=x, y=y)
  obj <- ggplot(datums, aes(x=x, y=y)) + 
    labs(title="Uniform Density of Dice rolls") +
    scale_x_discrete(name="x", limits=x) +
    geom_bar(stat = "identity", colour = 'black', fill = "lightblue")
  obj
}

do_dunif()
```

### Uniform Cumulative Distribution

Suppose that a die is rolled and $X$ is the number face up, what the probability that the value is greater than 1, 2, 3, 4, 5, or 6.

The plot shows there is a 100% prob over getting a value 1 or grea

```{r punif, echo=F, fig.width=6, fig.height=4, fig.align="center"}
do_punif <- function() {
  # PDF
  x <- seq(1, 6, 1)
  y <- punif(x, 1, 7, lower.tail=F)
  datums <- data.frame(x=x, y=y)
  obj <- ggplot(datums, aes(x=x, y=y)) + 
    labs(title="Uniform Cumulative Distribution") +
    scale_x_discrete(name=x, limits=x) +
    # geom_bar(stat = "identity", colour = 'black', fill = "lightblue")
    geom_line()

    obj
}

do_punif()
```
